---
title: "StabMap: Mosaic single cell data integration using non-overlapping features"
author:
  - Shila Ghazanfar
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: StabMap
vignette: |
  %\VignetteIndexEntry{Mosaic single cell data integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE
)
```


```{r setup, message=FALSE, warning=FALSE}
library(StabMap)
library(SingleCellMultiModal)
library(scran)
```

```{r}
set.seed(2021)
```

# Introduction

StabMap is a technique for performing mosaic single cell data integration. 

In this vignette we will elaborate on how these steps are implemented in the 
`StabMap` package.

# Load data

```{r}
mae <- scMultiome("pbmc_10x", mode = "*", dry.run = FALSE, format = "MTX")
```

Perform some exploration of this data.

```{r}
mae

upsetSamples(mae)

head(colData(mae))

dim(experiments(mae)[["rna"]])

names(experiments(mae))
```

Normalise and select features for the RNA modality.

```{r}
sce.rna <- experiments(mae)[["rna"]]

# Normalisation
sce.rna <- logNormCounts(sce.rna)

# Feature selection
decomp <- modelGeneVar(sce.rna)
hvgs <- rownames(decomp)[decomp$mean>0.01 & decomp$p.value <= 0.05]

length(hvgs)

sce.rna <- sce.rna[hvgs,]
```

Normalise and select features for the ATAC modality.

```{r}
dim(experiments(mae)[["atac"]])

sce.atac <- experiments(mae)[["atac"]]

# Normalise
sce.atac <- logNormCounts(sce.atac)

# Feature selection using highly variable peaks
# And adding matching peaks to genes
decomp <- modelGeneVar(sce.atac)
hvgs <- rownames(decomp)[decomp$mean>0.25
                         & decomp$p.value <= 0.05]
length(hvgs)

sce.atac <- sce.atac[hvgs,]
```

Create a composite full data matrix by concatenating.

```{r}
logcounts_all = rbind(logcounts(sce.rna), logcounts(sce.atac))
dim(logcounts_all)

assayType = ifelse(rownames(logcounts_all) %in% rownames(sce.rna),
                   "rna", "atac")
table(assayType)
```

# Mosaic data integration with StabMap

We will simulate a situation where half of the cells correspond to the Multiome
modality, and half of the cells correspond to the RNA modality. Our goal is
to then generate a joint embedding of the cells using all data, and to impute 
the missing ATAC values from the RNA modality cells.

```{r}
dataType = setNames(sample(c("RNA", "Multiome"), ncol(logcounts_all),
                           prob = c(0.5,0.5), replace = TRUE),
                    colnames(logcounts_all))
table(dataType)

assay_list = list(
  RNA = logcounts_all[assayType %in% c("rna"), dataType %in% c("RNA")],
  Multiome = logcounts_all[assayType %in% c("rna", "atac"), dataType %in% c("Multiome")]
)

lapply(assay_list, dim)
lapply(assay_list, class)
```

Examine the shared features between the two datasets using `mosaicDataUpSet()`.

```{r}
mosaicDataUpSet(assay_list, plot = FALSE)
```

From this we note that there are shared features between the RNA and Multiome
datasets, but there are many features that are observed only in the Multiome
dataset and not the RNA - as we had constructed.

We can understand the `mosaicDataTopology()` of these datasets, which
generates an `igraph` object, which can be inspected and plotted.

```{r}
mdt = mosaicDataTopology(assay_list)
mdt
plot(mdt)
```

From this we note that the datasets RNA and Multiome share at least some 
features. StabMap requires that the mosaic data topology network be connected,
that is, that there should be a path between every pair of nodes in the network.

We generate a common joint embedding for these data using StabMap. Since the 
Multiome data contains all features, we treat this as the reference dataset.
Since we already examined the mosaic data topology, we set `plot = FALSE`.

```{r}
stab = stabMap(assay_list,
               reference_list = c("Multiome"),
               plot = FALSE)
dim(stab)
stab[1:5,1:5]
```

We can reduce the dimension further using non-linear approaches such as UMAP.

```{r}
stab_umap = calculateUMAP(t(stab))
dim(stab_umap)

plot(stab_umap, pch = 16, cex = 0.3, col = factor(dataType[rownames(stab)]))
```

Here we see that the RNA and Multiome cells are fairly well-mixed.

# Data imputation after StabMap

Given the joint embedding, we can predict the missing ATAC values using
`imputeEmbedding()`. We provide the data list, the joint embedding as output 
from `stabMap()`. We set the Multiome cells as reference and the RNA cells as
query. This is useful for downstream visualisation or further interpretation.

```{r}
imp = imputeEmbedding(
  assay_list,
  stab,
  reference = colnames(assay_list[["Multiome"]]),
  query = colnames(assay_list[["RNA"]]))

class(imp)
names(imp)
lapply(imp, dim)
imp[["Multiome"]][1:5,1:5]
```

# Disjoint mosaic data integration with StabMap

StabMap is a flexible framework for mosaic data integration, and can still 
integrate data even when there are pairs of datasets that share no features at 
all. So long as there is a path connecting the datasets along the mosaic data 
topology (and the underlying assumption that the shared features along these
paths contain information), then we can extract meaningful joint embeddings. To
demonstrate this, we will simulate three data sources.

```{r}
dataTypeDisjoint = setNames(sample(c("RNA", "Multiome", "ATAC"), ncol(logcounts_all),
                                   prob = c(0.3,0.3, 0.3), replace = TRUE),
                            colnames(logcounts_all))
table(dataTypeDisjoint)

assay_list_disjoint = list(
  RNA = logcounts_all[assayType %in% c("rna"), dataTypeDisjoint %in% c("RNA")],
  Multiome = logcounts_all[assayType %in% c("rna", "atac"), dataTypeDisjoint %in% c("Multiome")],
  ATAC = logcounts_all[assayType %in% c("atac"), dataTypeDisjoint %in% c("ATAC")]
)

lapply(assay_list_disjoint, dim)
lapply(assay_list_disjoint, class)
```

Using `mosaicDataUpSet()`, we note that there are no shared features between 
the ATAC and RNA datasets. We might be able to match features by extracting
genomic positions and making the "central dogma assumption", that is, that the
peaks associated with a genomic position overlapping a gene should correspond to
positive gene expression for that gene. However, we need not make this 
assumption for the data integration to be performed.

```{r}
mosaicDataUpSet(assay_list_disjoint, plot = FALSE)
```

We can understand the `mosaicDataTopology()` of these datasets, which
generates an `igraph` object, which can be inspected and plotted.

```{r}
mdt_disjoint = mosaicDataTopology(assay_list_disjoint)
mdt_disjoint
plot(mdt_disjoint)
```

StabMap only requires that the mosaic data topology network be connected,
that is, that there should be a path between every pair of nodes in the network.
Since there is a path between RNA and ATAC (via Multiome), we can proceed.

We now generate a common joint embedding for these data using StabMap. Since the 
Multiome data contains all features, we again treat this as the reference 
dataset. Since we already examined the mosaic data topology, we set 
`plot = FALSE`.

```{r}
stab_disjoint = stabMap(assay_list_disjoint,
                        reference_list = c("Multiome"),
                        plot = FALSE)
dim(stab_disjoint)
stab_disjoint[1:5,1:5]
```

We can reduce the dimension further using non-linear approaches such as UMAP.

```{r}
stab_disjoint_umap = calculateUMAP(t(stab_disjoint))
dim(stab_disjoint_umap)

plot(stab_disjoint_umap, pch = 16, cex = 0.3,
     col = factor(dataTypeDisjoint[rownames(stab_disjoint)]))
```

Here we see that the RNA, ATAC and Multiome cells are fairly well-mixed.

Colouring the cells by their original cell type, we can also see that the 
mosaic data integration is meaningful.

```{r}
cellType = setNames(mae$celltype, colnames(mae[[1]]))

plot(stab_disjoint_umap, pch = 16, cex = 0.3,
     col = factor(cellType[rownames(stab_disjoint)]))
```

<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>
